---
title: "Data Tidying"
author: "Carolyn Kurle"
date: "January 16, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
library(dplyr)
library(tidyr)
```

To call a function from a specific package if the language for that function is the same between packages, type this: 'package_name::function_name(...)'

# Data Cleaning

Read in data File
```{r}
catch_df <-read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
                        method = "libcurl"),
                    stringsAsFactors = FALSE)


head(catch_df)
```
cntrl shift m is the shortcut for the pipe operator which looks like this: %<%
%>% 

We are now going to restructure the catch data into a tidy format. We created catch cleaned and moved some of the data from catch df into catch cleaned. So, the command below (catch_cleaned <- catch_df %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)) reads like this in English: catch_cleaned gets the values from catch df and then selects (pipes) certain columns from catch df to create a new data frame
  
  That removed the marginal sum and notes column

Next, we're going to gather the pertinent data into the key. We're going to gather everything except don't gather year, don't gather region.

This moved it from wide to long format and into the tidy version that we created when we did the hands on exercise with Matt. Now, the data are in a tall data table format that is what we want for analysis. We changed the name from Catch cleaned to catch long because the catch data is now in the long format data frame and it's not quite all the way clean. Jeanette wanted to save the name catch cleaned for the end.

```{r}

catch_long <- catch_df %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) %>% 
  gather(key = "species", value = "catch", -Year, -Region)


head(catch_long)
```


The catch data are listed in thousands of fish. So it would make sense to show this. 
First we're calling the data from catch long and then renaming catch to catch_thousands.

You can create new columns in your data using mutate. We wanted to change our numbers to numbers in the thousands and add that column. We got an error indicating that there's a word or nonnumberic in the catch column that can't be multiplied by 1000. We have to find that. We have to create a new block, then find that mistake.

We did test_catch to replace non-integers with NA. But that's not great. We assign a value to i which when you have na in test catch that equals true. That brings up our one error. We tried to force NA, but R didn't like that and said NAs were introduced by coercion. We then used the i format thing in the code below.

The hash tags are added in the below code to avoid R trying to run that problematic code (mutate)that we'll fix later.

To fix that error, we entered another mutate and we entered mutate(catch = ifelse). This is a conditional element selection. If the ifelse comes back as true, we want to have it replace it with a 1, or leave it if it's not. Then you have to turn that "1" into an integer which is the next mutate line. Before that, it was "1" as a letter. Which is weird. But now it's a number and not a character.

Then we can add the mutate line that turns the catch thousands into numbers that are actually multiplied by 1000.

Comments entered by Jeanette to address this were: eroneous value due to OCR issue = change "I" to one. Create catch column in correct units (thousands)

```{r}
catch_cleaned <- catch_long %>% 
  rename(catch_thousands = catch) %>%
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands)) %>% 
  mutate(catch_thousands = as.integer(catch_thousands)) %>% 
  mutate(catch = catch_thousands * 1000)

head(catch_cleaned)
```
```{r, eval = F, echo = F}

# I used this bit of code to find the bad value in the data set. Eval = F means that it won't evaluate this statement every single time and it won't show my code every single time I run this. But it will stay here as something instructional for me.

test_catch <- as.integer(catch_cleaned$catch_thousands)

i <- which(is.na(test_catch) == T)

catch_cleaned[i, ]
```
This allowed us to find the line (401) where our error was. This is an "l" but it should be a "1". Looking at the original data notes, we can see that that was supposed to be a 1 and we can go back and change it manually.

# Split-Apply-Combine

Calculate total catch by region

We want to group the data by region, then create a summarize column of the fish per region. We're creating a new data frame called catch_total. YOu could also group by year. I added that after region to do that as well. You can add multiple grouping variables separated by commas after the group by command.

We also added the number of observations per region per year. It's 5 in all cases here since there are only 5 species of salmon measured for each year and region.

```{r}
catch_total <- catch_cleaned %>% 
  group_by(Region, Year) %>% 
  summarize(catch_region = sum(catch),
            n_obs = n())

head(catch_total)
```

Now, I'm going to play with this group by function to think of other ways to group the data.

In this example, I decided I wanted to figure out the mean catch per species per year. I call the new data frame Catch_by_species_year and tell it to get it's data from the catch_cleaned frame. Then I "pipe" those data into a group by function that will group data by species and year then pipe those data into a summarize function that takes creates a column called catch_mean and that takes the mean of the column catch from the original catch_cleaned data.

```{r}
Catch_by_species_year <- catch_cleaned %>% 
  #the above means I've created a new frame that I've decided to call it Catch_by_species_year and that that new frame will use data from catch cleaned. then I entered the pipe function symbols that will tell it to pipe that outcome into the next command#
  group_by(species, Year) %>%
  
  #then I entered the group by function to tell it to group all the data by species and year. Then I do the pipe function again.#
  
  summarize(catch_mean = mean(catch))

#then I tell it to summarize those data into a new column in my new data from and it's going to be catch_mean and it's going to use the catch data from catch_cleaned for calculating those means#
  
head(Catch_by_species_year)
```

Now we're going to filter for chinook salmon only. NOte - that head command means it's going to show you the head of the table (the first 6 rows). Tail would show you the last few rows.

In the next gray box, we filtered for both chinook and chum

```{r}
catch_chinook <- catch_cleaned %>% 
  filter(species == "Chinook")

head(catch_chinook)
```
```{r}
catch_chinook_Chum <- catch_cleaned %>% 
  filter(species == "Chinook" |  species == "Chum")

#| means "or". You're telling it to filter for chinook or chum. Use the & if you want it to filter by 2 things at the same time. 
head(catch_chinook_Chum)
```

We could filter by a region and year.

```{r}
catch_chinook_region_year <- catch_cleaned %>% 
  filter(species == "Chinook" & Region == "SSE" & Year > 1990)

head(catch_chinook_region_year)
```
 
 Can also change how your sort or filter sorts. Ascending or descending.


# Joins


Now we are going to learn how to use joins within dplyr. We are going to use a left join to join the region data with the catch data. The region data is another data frame from our example. The left join is most common. Your left data frame is the one that is going to hold all of our data that we're only going to add a little bit of data to. The second is the right frame. In this case, our catch data is the left frame and the region data is the right.

We have to bring in our region data:


```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"),
                        stringsAsFactors = FALSE)

head(region_defs)
```

We're going to clean up the data a bit.

```{r}
region_clean <- region_defs %>% 
  select(code, mgmtArea)

head(region_clean)
```

Now we're going to join the two data frames. Call it something different so it will be its own frame.

We got an error as the code column titles weren't the same. Code vs. Region.

To get passed this we'll tell it to do the left join and then in the left frame it's called region and in the right frame it's called code. Now it knows those are OK. by the way, the c stands for concantinate.


```{r}
catch_join <- left_join(catch_cleaned, region_clean,
                        by = c("Region" = "code"))

head(catch_join)
```

You could also just rename the columns the same thing back at the beginning. That would have looked like this: 

region_clean <- region_defs %>% 
  select(code, mgmtArea)
  rename (Region = code)

head(region_clean)

When do you quote column names and when do you not? Dplyr will take quoted or non-quoted. Probably when in doubt, just use the quotations marks.  